Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-01-25 ms.tex
random

This paper is organized as followed.  In $\S\ref{sec:model}$ we
describe our model, including the sample of galaxies used in our
analysis ($\S\ref{sec:sample}$) and our numerical procedure for
calculating the steady-state hydrodynamic profile of the CNM
($\S\ref{sec:hydro}$).  In $\S\ref{sec:results}$ we describe our
results.  In $\S\ref{sec:applications}$ we describe applicaiton of our
results, including a comparison of our gas profiles to those inferred
from {\it Chandra} observations and to the measured $L_X-\Mbh$
relationship.  In $\S\ref{sec:discussion}$ we discuss our results.  In
$\S\ref{sec:summary}$ we summarize our results.
\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to measure the radial surface brightness profiles for hundreds of nearby early type galaxies. The measured brightness profile is generally well-fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law slope, $\Gamma$, to an outer power law slope, $\beta$, at a break radius, $\rb$.  Assuming the stellar population is spherically symmetric, this corresponds to a stellar density $\rhostar \propto r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for $r \gg \rb$.  To good approximation
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star(\rsoi) \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star(r_b) \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star(\rsoi)$ is the stellar density at the radius of the black hole sphere of influence (soi), 
\be
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\ee
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$ relationship of \citet{Gultekin+09},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
The use of equation (\ref{eq:Msigma}) is questionable for low mass black holes (e.g., \citealt{Greene&Ho07}), but our results are not sensitive to this assumption.

A galaxy model is fully specified by four parameters: $\Mbh$,
$\Gamma$, $r_b$, and $\beta$.  We compute models for three different
black hole masses, $\Mbh = 10^6$, $10^7$, $10^8 \Msun$.  The
distribution of $\Gamma$ in the \citet{LauerFaber+:2007a} sample is
bimodal, with a concentration of ``core" galaxies with $\Gamma \lsim
0.3$ and a concentration of ``cusp" galaxies with $\Gamma \gsim 0.7$.
We bracket these possibilities by considering models with $\Gamma=0.8$
and $\Gamma=0.1$.  Since our results are most sensitive to the
properties of the gas flow on radial scales $\ll r_b$, we fix $\beta =
2$ (a typical value). In any case, the flow inside $\rb$ is
insensitive the outer power law slope of the stellar density
profile. However, the presence of $\rb$ is necessary for to obtain a
converged steady state for some regions of our parameter space.  We
consider four different values of the $\rb$: 50 pc, 100 pc, 200 pc,
and 400 pc. These choices are motivated by the observed break radii in
the \citet{LauerFaber+:2007a} sample. For the core galaxies
($\Gamma \le$0.3) in this sample the break radius follows the scaling
relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with scatter of
approximately one dex. Most the cusp ($\Gamma \ge0.5$) galaxies have
$\rb$ between 100 and 1000 pc, and there is not trend with
$\Mbh$. {\bf AG: In practice when it is clear the rb will not have a
  significant effect on rs it may not be necessary to do runs. For
  instance for Mbh=$10^8$ Msun, gamma=0.8, and vw0=600 km/s the
  stagnation radius changes by only 8\%. So for cusp galaxies with
  vw0$\gsim$600 km/s and Mbh$\lsim 10^8$ Msun it seems like it is a
  safe bet to neglect the effect of the break radius on the
  solution. For low heating rates it my be difficult to converged
  solution for large values of rb=200 and 400 pc.}
% However, the presence of $\rb$ is necessary for to obtain a
% converged steady state for some regions of our parameter space. We
% adopt $\rb=100$ pc for cusp galaxies ($\Gamma=0.8$), while for core
% galaxy models we employ the scaling relation $\rb = $, motivated again
% by the \citet{LauerFaber+:2007a} sample. ({\bf BDM: fill in!}).
% Though simplistic, our results are not sensitive to our assumption
% about the value of $\rb$.

@@

The source term in the continuity equation (eq.~[\ref{eq:drhodt}]),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$
years.  To good approximation $\eta\simeq 0.02 (t_{\star}/t_h)^{-1.3}$ at time $t_{\star}$ following a star burst (e.g., \citealt{Ciotti+91}),\footnote{\citet{Ciotti+91} give the overall mass return rate from evolved stars as a function of B-band luminosity.  Although we give the mass return rate per unit volume, our expressions are equivalent.} although $\eta$ can be higher if star formation is ongoing (Fig X).

Source terms $\propto q$ also appear in the momentum and entropy
equations (eqs.~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) because isotropic
mass injection in the SMBH rest frame represents a source of momentum
and energy relative to the mean flow.  Physically, these result from
the mismatch between the properties of virialized gas injected locally
by stellar winds and the mean properties of the background flow.  The
term $\propto -q c_{s}^{2}$ acts as an important agent in stabilizing the flow to runaway cooling ($\S\ref{sec:instability}$).

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy
equation accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term accounts for the
minimum shock heating of the stellar winds due to the random movement
of stars with respect to each other at the characteristic velocity
$\sim \sigma = \sqrt{GM_{\rm enc}/r}$, as is present even for the low
wind velocities $\ll \sigma$ that characterize AGB mass loss in old
stellar populations.  The second term, $v_{w}^{2}$, parameterizes
the contribution of additional sources of heating, such as faster main
sequence or Wolf-Rayet winds, millisecond pulsars, supernovae, AGN
feedback, etc.  As discussed in $\S\ref{sec:heating}$, the value of
$v_{w}$ is uncertain and will in general depend on the SMBH mass and
the age of the stellar population.

To isolate the physics of interest, our baseline calculations neglect two potentially important effects: heat conduction and radiative cooling.  Heat transport by electron conductivity is included with an additional term to the right hand side of the entropy equation of the form  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T), \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering becomes comparable to the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field perpendicular to the flow could in principle suppress the
conductivity.  However, the radially-decreasing temperature profiles
of interest are susceptible to the magneto-thermal instability
(\citealt{Balbus01}), the non-linear evolution of which is believed to
result in a radially-directed field structure
(\citealt{Parrish&Stone07}).

Radiative cooling is also not included in our baseline calculations.  Across much of
the parameter space radiative cooling is subdominant compared to other
sources of cooling.  Radiative cooling cannot be neglected when $\vwO$
is low (and/or $\eta$ is high), but once important it leads to thermal
instabilities (e.g.~\citealt{McCourt+12}) which our 1D
time-independent model cannot account for self-consistently anyways.
Indeed, one of the main goals of this work is to use our analytic and
numerical solutions neglecting radiation to ascertain over what
physical conditions cooling instabilities are likely to develop
($\S\ref{sec:cooling}$).

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability
(\citealt{Brandenburg:2003a}).  We solve the equations for different
choices of $v_{w} = 300, 600, 1200$ km s$^{-1}$ spanning a
physically plausible range.  Although we are interested in the
steady-state intflow/outflow structure, the time-dependent equations
are solved to avoid numerical issues that arise near the critical
sonic points, of which there are up to two because of the supersonic
inflow on small radial scales and a possible supersonic outflow on
large scales.

The accuracy of our numerical solutions is assessed by checking the
level of conservation of mass across the grid, as well as the integral
constraint on the energy.  Mass and energy are conserved to $\lesssim
40$ per cent for all solutions presented in this paper.  {\bf BDM:
  more details...}

>>>>

This paper is organized as followed.  In $\S\ref{sec:model}$ we
describe our model, including the sample of galaxies used in our
analysis ($\S\ref{sec:sample}$) and our numerical procedure for
calculating the steady-state hydrodynamic profile of the CNM
($\S\ref{sec:hydro}$).  In $\S\ref{sec:results}$ we describe our
results.  In $\S\ref{sec:applications}$ we describe applicaiton of our
results, including a comparison of our gas profiles to those inferred
from {\it Chandra} observations and to the measured $L_X-\Mbh$
relationship.  In $\S\ref{sec:discussion}$ we discuss our results.  In
$\S\ref{sec:summary}$ we summarize our results.
\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to measure the radial surface brightness profiles for hundreds of nearby early type galaxies. The measured brightness profile is generally well-fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law slope, $\Gamma$, to an outer power law slope, $\beta$, at a break radius, $\rb$.  Assuming the stellar population is spherically symmetric, this corresponds to a stellar density $\rhostar \propto r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for $r \gg \rb$.  To good approximation
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star(\rsoi) \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star(r_b) \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star(\rsoi)$ is the stellar density at the radius of the black hole sphere of influence (soi), 
\be
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\ee
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$ relationship of \citet{Gultekin+09},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
The use of equation (\ref{eq:Msigma}) is questionable for low mass black holes (e.g., \citealt{Greene&Ho07}), but our results are not overly sensitive to this assumption.

A galaxy model is fully specified by four parameters: $\Mbh$,
$\Gamma$, $r_b$, and $\beta$.  We compute models for three different
black hole masses, $\Mbh = 10^6$, $10^7$, $10^8 \Msun$.  The
distribution of $\Gamma$ in the \citet{LauerFaber+:2007a} sample is
bimodal, with a concentration of ``core" galaxies with $\Gamma \lsim
0.3$ and a concentration of ``cusp" galaxies with $\Gamma \gsim 0.7$.
We bracket these possibilities by considering models with $\Gamma=0.8$
and $\Gamma=0.1$.  We fix $\beta = 2$ (a typical value) because our results are most sensitive to the
properties of the gas flow on radial scales $\ll r_b$ and do not depend sensitively on the outer power law slope of the stellar density profile. However, the presence of $\rb$ is necessary for to obtain a converged steady state for some regions of our parameter space.  Depending on the model, we
consider four different values of the $\rb$: 50 pc, 100 pc, 200 pc,
and 400 pc, a range motivated by the observed break radii in
the \citet{LauerFaber+:2007a} sample. For the core galaxies ($\Gamma \le$0.3) in this sample the break radius follows the scaling relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with scatter of
approximately one dex. Most the cusp ($\Gamma \ge0.5$) galaxies have
$\rb$ between 100 and 1000 pc, without a clear trend with $\Mbh$. {\bf AG: In practice when it is clear the rb will not have a
  significant effect on rs it may not be necessary to do runs. For
  instance for Mbh=$10^8$ Msun, gamma=0.8, and vw0=600 km/s the
  stagnation radius changes by only 8\%. So for cusp galaxies with
  vw0$\gsim$600 km/s and Mbh$\lsim 10^8$ Msun it seems like it is a
  safe bet to neglect the effect of the break radius on the
  solution. For low heating rates it my be difficult to converged
  solution for large values of rb=200 and 400 pc.}
% However, the presence of $\rb$ is necessary for to obtain a
% converged steady state for some regions of our parameter space. We
% adopt $\rb=100$ pc for cusp galaxies ($\Gamma=0.8$), while for core
% galaxy models we employ the scaling relation $\rb = $, motivated again
% by the \citet{LauerFaber+:2007a} sample. ({\bf BDM: fill in!}).
% Though simplistic, our results are not sensitive to our assumption
% about the value of $\rb$.

@@

The source term in the continuity equation (eq.~[\ref{eq:drhodt}]),
\begin{align}
  q=\frac{\eta \rhostar}{\th},
\label{eq:q}
\end{align}
represents mass input from stellar winds, which we parameterize in
terms of the fraction $\eta$ of the stellar density $\rhostar$ being
recycled into gas on the Hubble time $\th = 1.4 \times 10^{10}$
years.  To good approximation $\eta\simeq 0.02 (t_{\star}/t_h)^{-1.3}$ at time $t_{\star}$ following a star burst (e.g., \citealt{Ciotti+91})\footnote{\citet{Ciotti+91} give the overall mass return rate from evolved stars as a function of B-band luminosity.  Although we give the mass return rate per unit volume, our expressions are equivalent.}, although we will show later that $\eta$ can be higher if star formation is ongoing (Fig.~\ref{fig:eta}).

Source terms $\propto q$ also appear in the momentum and entropy
equations (eqs.~[\ref{eq:dvdt}] and [\ref{eq:dsdt}]) because isotropic
mass injection in the SMBH rest frame represents a source of momentum
and energy relative to the mean flow.  Physically, these result from
the mismatch between the properties of virialized gas injected locally
by stellar winds and the mean properties of the background flow.  The
term $\propto -q c_{s}^{2}$ is important because it acts to potentially stabilize the flow to runaway cooling ($\S\ref{sec:instability}$).

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy
equation accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random movement
of stars with respect to one another at the characteristic velocity
$\sim \sigma = \sqrt{GM_{\rm enc}/r}$.  This heating is present even for the low
wind velocities $\ll \sigma$ that characterize AGB mass loss in old
stellar populations.  The second term, $v_{w}^{2}$, parameterizes
the contribution of additional sources of heating, such as faster main
sequence or Wolf-Rayet stellar winds, spin-down energy from millisecond pulsars, supernovae, AGN
feedback, etc.  As discussed in $\S\ref{sec:heating}$, the value of
$v_{w}$ is uncertain and will in general depend on the SMBH mass and
the age of the stellar population.

To isolate the physics of interest, our baseline calculations neglect two potentially important effects: heat conduction and radiative cooling.  Heat transport by electron conductivity results in an an additional term on the right hand side of the entropy equation of the form  
\be \dot{q}_{\rm cond} = \nabla\cdot(\kappa \nabla T), \ee
where $\kappa_{\rm
  eff}=\kappa_{\rm spitz}/(1+\psi)$ (\citealt{DaltonBalbus:1993a}) is
the conducitivity and $\kappa_{\rm spitz} = \kappa_0 T^{5/2}$ is the
classical \citet{Spitzer62} value ($\kappa_0\simeq 2\times 10^{-6}$ in
cgs units).  The flux limitor $\psi = \kappa_{\rm spitz} \nabla
T/(5\phi \rho c_s^3)$ saturates the conductive flux if the mean free
path for electron coulomb scattering becomes comparable to the
temperature length scale, where $c_s \equiv (kT/\mu m_p)^{1/2}$ is the
isothermal sound speed and $\phi \lesssim 1$ is an uncertain
dimensionless constant (we adopt $\phi = 0.1$).  Even a weak magnetic
field perpendicular to the flow could in principle suppress the
conductivity by greatly reducing the electron mean free path.  However, for radially-decreasing temperature profiles
of interest, the flow is susceptible to the magneto-thermal instability (\citealt{Balbus01}), the non-linear evolution of which is believed to result in a radially-directed field geometry (\citealt{Parrish&Stone07}).

Radiative cooling is also not included in our baseline calculations.  Across much of the parameter space radiative cooling is subdominant compared to other sources of cooling.  Radiative cooling becomes important when $\vwO$ is low and/or $\eta$ is high, but once important it leads to thermal instabilities (e.g.~\citealt{McCourt+12}) which our 1D
time-independent model cannot account for self-consistently anyways.  Indeed, one of the main goals of this work is to use our analytic and numerical solutions neglecting radiation to ascertain over what range of physical conditions cooling instabilities will develop ($\S\ref{sec:cooling}$).

Equations (\ref{eq:drhodt})-(\ref{eq:dsdt}) are solved using a sixth
order finite difference scheme with a third order Runge-Kutta scheme
for time integration and artificial viscosity terms in the velocity
and entropy equations for numerical stability
(\citealt{Brandenburg:2003a}).  We assume different
choices of $v_{w} = 300, 600, 1200$ km s$^{-1}$ spanning a
physically plausible range of heating rates.  Although we are interested in the
steady-state inflow/outflow structure (assuming it exists), the time-dependent equations are solved to avoid numerical issues that arise near the critical sonic points, of which there are up to two because of the supersonic
inflow on small radial scales and a possible supersonic outflow on large scales.

We assess the accuracy of our numerical solutions by checking that mass is conserved across the grid, in addition to the integral constraint on the energy (Bernoulli integral).  Mass and energy are conserved to $\lesssim
40$ per cent for all solutions presented in this paper.  {\bf BDM:
  more details...}
